Statically Typed Web Programming using Scala

### 2012-12-18 papers/ecoop2013/repairing.tex
Commit rapide

\begin{abstract}
Rich Internet Applications involve even more code on the client-side, dealing with DOM manipulation, event handling
and asynchronous calls to the server. Writing and maintaining large JavaScript code bases is challenging because this
language has no static typing and no module system. We investigate further the definition of JavaScript as a Scala
compiled embedded DSL using Lightweight Modular Staging. We define safe and expressive abstractions for DOM
manipulation, event handling and asynchronous programming that translate to efficient JavaScript code using native
APIs. We use both type-level information and staging to introduce high-level abstractions with few or no overhead on
the generated JavaScript code.
\end{abstract}

@@

This paper continues their work and presents several safe and expressive DSL constructs for RIAs development. We
increase the expressivity of JavaScript native APIs with no runtime overhead and expose them through statically typed
APIs. We make asynchronous programming simpler by raising asynchronous computations to first-class citizens. We
leverage the Scala type system to bring type-directed language features such as \emph{ad hoc} polymorphism and type
coercion to the client-side.

@@

Dynamic typing may give programmers more flexibility but can make it harder for little scripts to grow into mature
and robust code bases and to perform refactorings. For instance, in the case of RIAs, a mispell in an event name may
break the program behavior without anyone getting a sensible error message.

@@

\subsubsection{Sharing type safe code between clients and servers is hard to achieve}

@@

The following Scala listing generates code equivalent to listing \ref{event-js} and is completely type safe: if
the user misspells the event name or tries to use an undefined property on an event his program won’t compile.

@@

The \code{\~} type is just an alias to the \code{Args} type to get the infix notation. The \code{Args} type
aggregates the types of the operands (\code{Lhs} and \code{Rhs}) involved in a binary operation.

@@

We limited the DSL to two operations (\code{foreach} and \code{isEmpty}) for the sake of brevity. The
\code{OptionOps} trait defines the user facing DSL, it enriches \code{Rep[Option[_]]} values by providing an implicit
conversion to \code{OptionOpsCls} that defines the \code{isEmpty} and \code{foreach} methods that respectively tests
the emptiness of a value and derefences a value if not empty. These methods provide a convenient syntax to use
\code{Rep[Option[_]]} values and delegate to the \code{option_isEmpty} and \code{option_foreach} operations,
respectively, that are left abstract. The \code{OptionOpsExp} trait implements the \code{option_isEmpty} and
\code{option_foreach} methods. The \code{option_isEmpty} implementation simply returns a statement node of type
\code{OptionIsEmpty}, but the \code{option_foreach} implementation is more subtle because it has to track the
dependencies of its function parameter \code{f} (for instance, side-effects performed within the \code{f} function
must be reflected by the returned \code{OptionForeach} node).

@@

Our goal was to define, with a minimal effort, a language for RIAs development. The main challange was to provide
programming language constructs both safer and more expressive without abstraction penalty. We focused on the main
concerns of RIA development: event handling, asynchronous programming and DOM fragments definition, as well as on
more general programming concerns: \code{null} reference handling and type safety.

@@

The most popular JavaScript library, jQuery~\cite{Bibeault08_jQuery}, used by more than 40\% of the top million
sites\footnote{\href{http://trends.builtwith.com/javascript}{http://trends.builtwith.com/javascript}} aims to
simplify RIA development. It handles the \code{null} references problem by wrapping each query result in a container
so before each further method call it tests the emptiness of the container and applies effectively the operation only
if the container is not empty. It provides an expressive API but the emptiness checking involves a slight performance
penalty since each result is wrapped and then unwrapped for the subsequent method invocation. jQuery also has an API
to turn asynchronous computations into first-class citizens. However, jQuery is just a library and thus cannot bring
type safety to JavaScript programs.

@@

GWT~\cite{Chaganti07_GWT} compiles Java code to JavaScript. It exposes the browser’s APIs through type safe Java APIs
and simplifies DOM fragments definition. However, it still uses a callback-driven programming model for asynchronous
computations.

@@

Table \ref{comparison} gives a synthetic view of the comparison of our work with other approaches. Lines give
features desired to write large Web applications and colums list mainstream tools for Web programming. Cells show the
support level of each feature for each tool.

@@

>>>>

\begin{abstract}
Rich Internet Applications rely on ever more code on the client-side, 
typically written in JavaScript, dealing
with DOM manipulation, event handling and asynchronous calls to the 
server. Beyond well documented JavaScript programming intrinsic issues,
the hard wired distribution of an application business logic across 
clients and servers hampers its flexibility and maintainability. This 
paper presents a Scala compiled
embedded DSL for Web programming trying to address these shortcomings. 
Close to JavaScript, it provides type-safe
and expressive abstractions for DOM manipulation, event handling and
asynchronous programming. Code written using the DSL can be compiled to both
server and client sides. Built on top of the Lightweight Modular Staging
framework, this DSL uses both type-level information and staging to  introduce
high-level abstractions that translate to efficient JavaScript and Scala 
code using native APIs.
\end{abstract}

@@

Based on this work, this paper introduces several type-safe and expressive DSL constructs for RIAs development. This
DSL increases the expressivity of JavaScript native APIs with no runtime overhead and exposes them through statically
typed APIs. It makes asynchronous programming simpler by raising asynchronous computations to first-class citizens.
It leverages the Scala type system to bring type-directed language features such as \emph{ad hoc} polymorphism and
type coercion to the client-side. Finally, code written using the DSL can be compiled to both server and client
sides.

@@

Dynamic typing may give programmers more flexibility but can make it harder for little scripts to grow into mature
and robust code bases and to perform refactorings~\cite{Meijer04_StaticDynamic}. For instance, in the case of RIAs, a
mispell in an event name may break the program behavior without anyone getting a sensible error message.

@@

\subsubsection{Sharing type-safe code between clients and servers is hard to achieve}

@@

The following Scala listing generates code equivalent to listing \ref{event-js} and is completely type-safe: if
the user misspells the event name or tries to use an undefined property on an event his program won’t compile.

@@

The \code{\~} type is just an alias for the \code{Args} type to get the infix notation. The \code{Args} type
aggregates the types of the operands (\code{Lhs} and \code{Rhs}) involved in a binary operation.

@@

We limited the listing to two operations (\code{foreach} and \code{isEmpty}) for the sake of brevity. The
\code{OptionOps} trait defines the user facing DSL, it enriches \code{Rep[Option[_]]} values by providing an implicit
conversion to \code{OptionOpsCls} that defines the \code{isEmpty} and \code{foreach} methods that respectively tests
the emptiness of a value and derefences a value if not empty. These methods provide a convenient syntax to use
\code{Rep[Option[_]]} values and delegate to the \code{option_isEmpty} and \code{option_foreach} operations,
respectively, that are left abstract. The \code{OptionOpsExp} trait implements the \code{option_isEmpty} and
\code{option_foreach} methods. The \code{option_isEmpty} implementation simply returns a statement node of type
\code{OptionIsEmpty}, but the \code{option_foreach} implementation is more subtle because it has to track the
dependencies of its function parameter \code{f} (for instance, side-effects performed within the \code{f} function
must be reflected by the returned \code{OptionForeach} node).

@@

Our goal was to define, with a minimal effort, a language for RIAs development. The main challenge was to provide
programming language constructs that are both safer and more expressive without performance penalty. We focused on
the main concerns of RIA development: event handling, asynchronous programming and DOM fragments definition, as well
as on more general programming concerns: \code{null} reference handling and type safety.

@@

The most popular JavaScript library, jQuery~\cite{Bibeault08_jQuery}, used by more than 40\% of the top million
sites\footnote{\href{http://trends.builtwith.com/javascript}{http://trends.builtwith.com/javascript}} aims to
simplify RIA development. It handles the \code{null} references problem by wrapping each query result in a container
so before each further method call it tests the emptiness of the container and applies the operation effectively only
if the container is not empty. It provides an expressive API but the emptiness checking involves a slight performance
penalty since each result is wrapped and then unwrapped for the subsequent method invocation. jQuery also has an API
to turn asynchronous computations into first-class citizens. However, jQuery is just a library and thus cannot bring
type safety to JavaScript programs.

@@

GWT~\cite{Chaganti07_GWT} compiles Java code to JavaScript. It exposes the browser’s APIs through type-safe Java APIs
and simplifies DOM fragments definition. However, it still uses a callback-driven programming model for asynchronous
computations.

@@

Table \ref{comparison} gives a synthetic view of the comparison of our work with other mainstream approaches. Lines
give features desired to write large Web applications and colums list mainstream tools for Web programming. Cells
show the support level of each feature for each tool.

@@

Several model-driven methodologies for creating web applications have been proposed, including
\cite{schwabe1996systematic,lima2003modeling,ceri2000web,koch2001authoring,pastor2003oows,valverde2007mda,vdovjak2003engineering,kraus2007model,nunes2006rapid,brambilla2008designing,valderas2007transformational,van2006hera,Groenewegen08_WebDSL}.
Many of them follow the standard top-down MDE vision in providing abstractions and code generators, nevertheless
these cannot be easily customized or generate only application skeletons.
